CrypTool Script \\ Mathematics and Cryptography

### 2003-03-07 script/en/chapters/primes.tex
no message

\newpage
\section{Prime numbers}\hypertarget{Kapitel_2}{}
(Bernhard Esslinger, besslinger@web.de, May 1999, Updates Nov. 2000, December 2001)

@@

\subsection{What are prime numbers?}
\index{Prime number} \index{Number!prime}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

@@

\subsection{Prime numbers in mathematics}\label{primesinmath}

@@

\subsection{How many prime numbers are there?}

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid\footnote{Euclid,
a Greek mathematician of 4th and 3rd century before Christ. He worked at the
egyptian academy of Alexandria and wrote ``The Elements'', the most well known systematical textbook
of the Greek mathematics.} knew this in the
third century before Christ.
\begin{theorem}[Euclid]\label{thm-pz-euklid}\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid’s Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as following
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than any previously existing amount
of prime numbers.}
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

\subsection{The search for extremely large primes}

@@

@@

$$
\begin{array}{cl}
2^2 - 1 = 3 & \Rightarrow {\rm prime} \\
2^3 - 1 = 7 & \Rightarrow {\rm prime} \\
2^5 - 1 = 31    & \Rightarrow {\rm prime} \\
   \vdots \\
2^{11} - 1 = 2.047 = 23 \cdot 89    & \Rightarrow  {\rm NOT~prime} !
\end{array}
$$

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before \index{Lucas
  Edouard} Edouard Lucas (1842-1891) proved that this number is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson showed in 1903 which factors make up this composite number: $$ 2^{67} -1
=147, 573, 952, 588, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorisation (expression as a
product of factors) of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers \index{Mersenne!prime number} 
have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377; ~
6,972,593; \\13,466,917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known.

@@

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\vskip - 4pt \noindent
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\vskip - 4pt \noindent
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}
\subsubsection{M-37 -- January 1998}\index{Mersenne!prime number!M-37}
The 37th Mersenne prime,
$$ 2^{3.021.377} - 1. $$
was found in January 1998 and has 909,526
digits in the decimal system, which corresponds to 33 pages in the newspaper!

\subsubsection{M-38 -- June 1999}\index{Mersenne!prime number!M-38}
The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

\subsubsection{M-39 -- December 2001} \index{Mersenne!prime number!M-39}

The 39th Mersenne prime, called M-39, $$2^{13,466,917}-1,$$ was found at December 6, 2001 –- more exactly,
the verification of this number, found at November 14, 2001 by the Canadian student Michael Cameron,
was successfully finished. This number has about 4 million decimal digits (exactly 4,053,946 digits).
Trying only to print this number $(924947738006701322247758 \cdots 1130073855470256259071)$ in the Financial Times
would need around 200 pages.

\subsubsection{GIMPS}\index{GIMPS}

Discovering the 39th Mersenne prime the GIMPS project (Great Internet Mersenne Prime Search) founded in 1996
already discovered for the 5th time the greatest Mersenne number to be proven being prime.

@@

\subsubsection{EFF}\index{EFF}

This search is also spurred on by a competition started by the non-profit
organisation EFF (Electronic Frontier Foundation) using the means of an unknown
donator. The participants are rewarded with a total of 500,000 USD if they find
the longest prime number. In promoting this project, the unknown donator is not
looking for the quickest computer, but rather wants to draw people's attention
to the possibilities offered by {\em cooperative networking}

{\href{http://www.eff.org/coopawards/prime-release1.html}{\tt http://www.eff.org/coopawards/prime-release1.html}}.

@@

\subsection{Prime number tests}
\index{Prime number!test}
In order to implement secure encryption procedures we need extremely large prime
numbers (in the region of $2^{2,048}$, i.e. numbers with $600$ digits in the
decimal system!).

@@

However, if we know something about the {\em construction} of the number in
question, there are extremely highly developed procedures that are much quicker.
In the 17th century, Fermat wrote to Mersenne that he presumed that all numbers
of the form $$ F(n) = 2^{2^n} + 1 $$ are prime for all whole numbers $n$ greater
than or equal to $0$ (see below).

@@

Comment: Numbers that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$ but
are not prime are called \index{Pseudoprime number}\index{Number!pseudoprime} {\em pseudoprime numbers}. The first pseudoprime number (i.e. not a
prime) is $$ 341 = 11 \cdot 31. $$
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Number!Carmichael} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17. $$
A stronger test is provided by \index{Miller} \index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudoprime numbers}. Again, there are
strong pseudoprime numbers that are not primes, but this is much less often the
case than for (simple) pseudoprime numbers. The smallest strong pseudoprime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudoprime number up to $25 \cdot 10^9$, i.e. a number that passes the test and
yet is not a prime number.

@@

\subsection{The search for a formula for prime numbers}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

Ideas for this can be found at

{\href{http://www.utm.edu/research/primes/notes/faq/p_n.html}
 {\tt http://www.utm.edu/research/primes/notes/faq/p\_n.html}}.

Cross-reference:  \hyperlink{ntePrimzahl}{the table under \ref{s:ntePrimzahl}} contains the precise values for the $n$th
prime numbers for selected $ n.$

The following are some of the most common ideas for prime number formulae:
\begin{enumerate}
  \item Mersenne numbers  $f(n) = 2^n - 1$: \\
    As shown above, this formula seems to deliver relatively large prime numbers
but - as for $n=11$ [$f(n)=2,047$] - it is repeatedly the case that the result
is not prime.
  \item $F(k,n) = k \cdot 2^n \pm 1$  for $ n $ prime and $ k $ small primes:\\
   For this generalisation of the Mersenne numbers there are (for small $k$)
also extremely quick prime number tests (see \cite{Knuth1981}). This can be
performed in practice using software such as the PROTHS software from Yves
Gallot

({\href{http://www.prothsearch.net/index.html}{\tt
http://www.prothsearch.net/index.html}}).

  \item {\em Fermat numbers}\footnote{The Fermat prime numbers play a role in circel division.
As proven by Gauss\index{Gauss}, a regular $p$-edge
can only be constructed with the use of a pair of compasses and a ruler, when $p$ is a Fermat prime number.}
\index{Number!Fermat}\index{Fermat!number}  $F(n) = 2^{2^n} + 1$:\\
As mentioned above, Fermat wrote to Mersenne regarding this assumption.
Surprisingly he would have been able obtain a positive result using the negative
prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = 65,537 &  \mapsto {\rm ~prime}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = 4 ,294, 967, 297 = 641 \cdot
6,700,417 &  \mapsto {\rm ~NOT~prime} !
\end{array} 
$$
  \item Carmichael numbers: see above.
  \item Pseudoprime numbers: see above.
  \item Strong pseudoprime numbers: see above.
  \item Idea based on \hyperlink{euclid}{Euclid's proof} (infinite many prime
numbers) $p_1 \cdot p_2 \cdots p_n +1$:

@@

\subsection{Density and distribution of the primes}

@@

\subsection{Notes}

@@

\begin{itemize}
  \item The German mathematician \index{Goldbach Christian} Christian Goldbach
(1690-1764) conjectured:
    Every even natural number greater than 2 can be represented as the sum of
two prime numbers.
    Computers have verified\footnote{It is generally accepted today, that
the Goldbach Conjecture is true, i.~e.\ valid for all even natural numbers bigger than $2$.
In 1999, mathematician J\"org Richtstein from the computer sciences institute at the University
of Giessen, studied even numbers up to 400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact that despite all efforts, Goldbachs' conjecture has to date not been proven leads
one to believe that: since the pioneer work of the Austrian mathematician Kurt G\"odel \index{G\""odel Kurt} is well-known,
not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbachs' Conjecture was correct, despite that however,
proof will never be found. Conversely, that will presumably also remain unproven.}
the Goldbach conjecture for all even numbers up to
$4*10^{14}$ but no general proof has yet been found\footnote{The English
publisher {\em Faber} and the American publisher {\em Bloomsbury} issued in 2000
the 1992 published book ``Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than 250
year old puzzle. To
stimulate the sell rates the English and American publishers compete a prize of 1 million USD, 
if someone proves the conjecture -- to be published till 2004 by a well-known mathematical journal.\\
Susprisingly only British and American citizens are allowed to participate.\\
The theorem closest to Goldbach's conjecture yet was proven 1966 by Chen Jing-Run in a hard to understand fashion:
Each even integer bigger than $2$ is the sum of one prime and of the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.\\
Especially this conjecture makes transparent, that even today we do not
have fully understood the deeper connections between addition and
multiplication of natural numbers.}.
  \item The German \index{Riemann Bernhard} mathematician Bernhard Riemann
(1826-1866) put forward a formula for the distribution of primes that would
further improve the estimate. However, this has neither been proved nor
contradicted so far.
\end{itemize}

@@

\newpage
\subsubsection{Number of prime numbers in various intervals}\label{s:primhfk}
\hypertarget{primhfk}{}

@@

\newpage
\subsubsection{Indexing prime numbers ($n$-th prime number)} \hypertarget{ntePrimzahl}{}\label{s:ntePrimzahl}

@@

\newpage\begin{thebibliography}{99999}
\addcontentsline{toc}{subsection}{Bibliography}
\bibitem[Bartholome1996]{Bartholome1996}  \index{Bartholome 1996}     A.
Bartholomé, J. Rung, H. Kern, \\     Zahlentheorie f\"ur Einsteiger, Vieweg 1995,
2. Auflage 1996.

>>>>

\newpage
\section{Prime numbers}
\hypertarget{Kapitel_2}{}
(Bernhard Esslinger, besslinger@web.de, May 1999, Updates Nov. 2000, December 2001, February 2003)

@@

% --------------------------------------------------------------------------
\subsection{What are prime numbers?}
\index{Prime number} \index{Number!prime}
Prime numbers are whole, positive numbers greater than or equal to $2$ that can
only be divided by 1 and themselves. All other natural numbers greater than or
equal to $2$ can be formed by multiplying prime numbers.

@@

% --------------------------------------------------------------------------
\subsection{Prime numbers in mathematics}\label{primesinmath}

@@

% --------------------------------------------------------------------------
\subsection{How many prime numbers are there?}

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century before Christ. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century before Christ.
\begin{theorem}[Euclid]\label{thm-pz-euklid}\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid’s Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than any previously existing amount
of prime numbers.}
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

% --------------------------------------------------------------------------
\subsection{The search for extremely large primes}

@@

% --------------------------------------------------------------------------
\hypertarget{MersenneNumbers01}{}
\subsubsection{Special number types -- Mersenne numbers} \index{ Mersenne numbers}

@@

$$
\begin{array}{cl}
2^2 - 1 = 3 & \Rightarrow {\rm prime} \\
2^3 - 1 = 7 & \Rightarrow {\rm prime} \\
2^5 - 1 = 31    & \Rightarrow {\rm prime} \\
2^7 - 1 = 127    & \Rightarrow {\rm prime} \\
2^{11} - 1 = 2.047 = 23 \cdot 89    & \Rightarrow  {\rm NOT~prime} !
\end{array}
$$

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before \index{Lucas
  Edouard} Edouard Lucas (1842-1891) proved that this number is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson showed in 1903 which factors make up this composite number: $$ 2^{67} -1
=147, 573, 952, 588, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorisation 
\index{factorisation} (expression as a product of factors) 
\footnote{%
Using CrypTool\index{CrypTool} v1.3 you can factorize numbers in the 
following way: menu {\bf Indiv. Procedures \textbackslash RSA Demonstration 
\textbackslash Factorisation of a Number}.
}
of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers \index{Mersenne!prime number} 
have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377; ~
6,972,593; \\13,466,917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known. For the first 38 Mersenne prime numbers we know that this list is complete. The exponents until the 39th Mersenne prime numbers are not yet checked completely.

@@

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\vskip - 4pt \noindent
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\vskip - 4pt \noindent
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}

\vskip +15 pt
\paragraph{M-37 -- January 1998}\index{Mersenne!prime number!M-37}\mbox{}
% \\
The 37th Mersenne prime,
$$ 2^{3,021,377} - 1. $$
was found in January 1998 and has 909,526
digits in the decimal system, which corresponds to 33 pages in the newspaper!

\vskip +15 pt
\paragraph{M-38 -- June 1999}\index{Mersenne!prime number!M-38}\mbox{}
% \\
The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

\vskip +15 pt
\paragraph{M13466917 -- December 2001} \index{Mersenne!prime number!M-39}\mbox{}
% \\
This number was discovered as 39th Mersenne prime (and already called M-39, despite it is not proven yet, whether there do not exist further Mersenne prime numbers between M-38 und M13466917), $$2^{13,466,917}-1,$$ at December 6, 2001 –- more exactly, the verification of this number, found at November 14, 2001 by the Canadian student Michael Cameron, was successfully finished. This number has about 4 million decimal digits (exactly 4,053,946 digits).
Trying only to print this number $(924947738006701322247758 \cdots 1130073855470256259071)$ in the Financial Times would need around 200 pages.

\vskip +15 pt
\paragraph{GIMPS}\index{GIMPS}\mbox{}
% \\
Discovering the 39th Mersenne prime the GIMPS project (Great Internet Mersenne Prime Search) founded in 1996 already discovered for the 5th time the greatest Mersenne number to be proven being prime
({\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}).

@@

Currently the 4 largest known primes are Mersenne prime numbers. Only the 5th biggest known prime is another number type: it belongs to the so called  \hyperlink{generalizedFermatprimes}{generalized Fermat primes}.

% --------------------------------------------------------------------------
\vskip +15 pt
\subsubsection{EFF}\index{EFF}
This search is also spurred on by a competition started by the non-profit
organisation EFF (Electronic Frontier Foundation) using the means of an unknown
donator. The participants are rewarded with a total of 500,000 USD if they find
the longest prime number. In promoting this project, the unknown donator is not
looking for the quickest computer, but rather wants to draw people's attention
to the possibilities offered by {\em cooperative networking} \\
{\href{http://www.eff.org/coopawards/prime-release1.html}{\tt http://www.eff.org/coopawards/prime-release1.html}}

@@

% --------------------------------------------------------------------------
\subsection{Prime number tests}
\index{Prime number!test}
In order to implement secure encryption procedures we need extremely large prime
numbers (in the region of $2^{2,048}$, i.e. numbers with $600$ digits in the
decimal system!).

@@

However, if we know something about the {\em construction} of the number in
question, there are extremely highly developed procedures that are much quicker.

\hypertarget{FermatNumbers01}{}
In the 17th century, Fermat wrote to Mersenne that he presumed that all numbers
of the form $$ F(n) = 2^{2^n} + 1 $$ are prime for all whole numbers $n$ greater
than or equal to $0$ (\hyperlink{FermatNumbers02}{see below}) 
\index{Zahlen!Fermatzahlen}\index{Fermat!Fermatzahl}

@@

Comment: Numbers that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$ but
are not prime are called \index{Pseudo prime number}\index{Number!pseudo prime} {\em pseudo prime numbers}. The first pseudo prime number (i.e. not a
prime) is $$ 341 = 11 \cdot 31. $$
There are numbers that pass the Fermat test with all bases and yet are not
prime: these numbers are called \index{Number!Carmichael} {\em
Carmichael numbers}. The first of these is $$ 561 = 3 \cdot 11 \cdot 17. $$
A stronger test is provided by \index{Miller} \index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudo prime numbers}. Again, there are
strong pseudo prime numbers that are not primes, but this is much less often the
case than for (simple) pseudo prime numbers. The smallest strong pseudo prime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

% --------------------------------------------------------------------------
\vskip +20 pt
\subsection{Further special number types and the search for a formula for primes}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

Ideas for this can be found at \\
{\href{http://www.utm.edu/research/primes/notes/faq/p_n.html}
 {\tt http://www.utm.edu/research/primes/notes/faq/p\_n.html}}.

Cross-reference:  \hyperlink{ntePrimzahl}{the table under \ref{s:ntePrimzahl}} contains the precise values for the $n$th
prime numbers for selected $ n.$
\\

The following enumeration contains the most common ideas for ``prime number formulae'':
\begin{enumerate}

\item Mersenne numbers  $f(n) = 2^n - 1$: \\
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems to deliver 
    relatively large prime numbers but - as for $n=11$ [$f(n)=2,047$] - it is 
    repeatedly the case that the result is not prime.
    Today, all the Mersenne primes having less than 2,000,000 digits are known\\
    ({\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}).

\item $F(k,n) = k \cdot 2^n \pm 1$  for $ n $ prime and $ k $ small primes:\\
   For this generalisation of the Mersenne numbers there are (for small $k$)
   also extremely quick prime number tests (see \cite{Knuth1981}). This can be
   performed in practice using software such as the PROTHS software from Yves Gallot \\
   ({\href{http://www.prothsearch.net/index.html}
    {\tt http://www.prothsearch.net/index.html}}).

\vskip +15 pt
\hypertarget{FermatNumbers02}{}
\item {\em Fermat numbers}
\footnote{The Fermat prime numbers play a role in circle division.
As proven by Gauss\index{Gauss}, a regular $p$-edge
can only be constructed with the use of a pair of compasses and a ruler, when $p$ is a Fermat prime number.}
\index{Number!Fermat}\index{Fermat!number}  $F(n) = 2^{2^n} + 1$:\\
As mentioned \hyperlink{FermatNumbers01}{above}, Fermat wrote to Mersenne regarding his assumption, that all numbers of this type are primes.
Surprisingly he would have been able obtain a positive result using the negative
prime number test for $n=5$ based on his small theorem.
$$
\begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prime}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prime}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prime}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prime}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = 65,537 &  \mapsto {\rm ~prime}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = 4 ,294, 967, 297 = 641 \cdot
6,700,417 &  \mapsto {\rm ~NOT~prime} !
\end{array} 
$$

    Within the project ``Distributed Search for Fermat Number Dividers'' offered by
    Leonid Durman there is also progress in finding new monster primes
    ({\href{http://www.fermatsearch.org/}
       {\tt http://www.fermatsearch.org/}}).
    This website has conjectured webpages in Russian, Italian and German.

    The discovered factors can be compound integers or primes.

    At February 22, 2003 John Cosgrave discovered 
    \begin{itemize} 
     \item the largest composite Fermat number yet and
     \item the largest prime non-Mersenne prime yet with 645,817 digits.
    \end{itemize}

    The Fermat number
    $$ F[2,145,351] = 2^(2^{2,145,351}) + 1 $$ 
    is divisible by the prime
    $$ p = (3*2^{2,145,353} + 1) $$ \\
    This prime is the largest known prime generalized Mersenne number. 
    It was discovered only a few days after Michael Angel discovered the largest
    generalized Fermat number (\hyperlink{generalizedFermatprimes}{see below}).
    This moved GIMPS first prime discovery, M-35 = M1398269, into 7th place.

    This work was done using NewPGen from Paul Jobling's, PRP from George Woltman's, 
    Proth from Yves Gallot's programs and also the Proth-Gallot group at St. Patrick's 
    College, Dublin.

    More details are in\\
    {\href{http://www.fermatsearch.org/history/cosgrave_record.htm/}
       {\tt http://www.fermatsearch.org/history/cosgrave\_record.htm}}.

%    ({\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
%           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}).
% M.E. ist die Zahl auf seiner Webseite zu groß, da M-39 "nur" 4 Mio. Stellen hat !
%    Heute kennt man alle Fermatschen Primzahlen mit bis zu 2.000.000.000 ??? 
%    Dezimalstellen. \\

\vskip +15 pt
\hypertarget{generalizedFermatprimes}{}
\item Generalized Fermat numbers    $F(b,m) = b^{2^m} + 1$: \\
\index{Fermat numbers}
    Generalized Fermat numbers are more numerous than Mersenne numbers at equal size 
    and many of them are waiting to be discovered to fill the big gaps between the 
    Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the representation is not 
    limited to the base 2, can be tested at almost the same speed than a 
    Mersenne number.

    Yves Gallot wrote the program Proth.exe to investigate generalized Fermat numbers.

    Using this program at February 16, 2003 Michael Angel discovered the largest of them 
    till then with 628,808 digits, which so became the 5th largest known prime number:
    $$ b^{2^{17}} + 1  =  62722^{131,072} + 1. $$ 
    This moved GIMPS first prime discovery, M-35 = M1398269, into 6th place.

   ({\href{http://www.prothsearch.net/index.html}
    {\tt http://www.prothsearch.net/index.html}})

   ({\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
    {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}})

  \vskip +15 pt
  \item Carmichael numbers: see above.

  \item Pseudo prime numbers: see above.

  \item Strong pseudo prime numbers: see above.

  \item Idea based on \hyperlink{euclid}{Euclid's proof} (infinite many prime
numbers) $p_1 \cdot p_2 \cdots p_n +1$:

@@

% --------------------------------------------------------------------------
\subsection{Density and distribution of the primes}

@@

% --------------------------------------------------------------------------
\subsection{Notes}

@@

\begin{itemize}
  \item The German mathematician \index{Goldbach Christian} Christian Goldbach
(1690-1764) conjectured:
    Every even natural number greater than 2 can be represented as the sum of
two prime numbers.
    Computers have verified\footnote{It is generally accepted today, that
the Goldbach Conjecture is true, i.~e.\ valid for all even natural numbers bigger than $2$.
In 1999, mathematician J\"org Richstein from the computer sciences institute at the University
of Giessen, studied even numbers up to 400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact that despite all efforts, Goldbachs' conjecture has to date not been proven leads
one to believe that: since the pioneer work of the Austrian mathematician Kurt G\"odel \index{G\""odel Kurt} is well-known,
not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbachs' Conjecture was correct, despite that however,
proof will never be found. Conversely, that will presumably also remain unproven.}
the Goldbach conjecture for all even numbers up to
$4*10^{14}$ but no general proof has yet been found\footnote{The English
publisher {\em Faber} and the American publisher {\em Bloomsbury} issued in 2000
the 1992 published book ``Uncle Petros and Goldbach's Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than 250
year old puzzle. To
stimulate the sell rates the English and American publishers compete a prize of 1 million USD, 
if someone proves the conjecture -- to be published till 2004 by a well-known mathematical journal.\\
Surprisingly only British and American citizens are allowed to participate.\\
The theorem closest to Goldbach's conjecture yet was proven 1966 by Chen Jing-Run in a hard to understand fashion:
Each even integer bigger than $2$ is the sum of one prime and of the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.\\
Especially this conjecture makes transparent, that even today we do not
have fully understood the deeper connections between addition and
multiplication of natural numbers.}.
  \item The German \index{Riemann Bernhard} mathematician Bernhard Riemann
(1826-1866) put forward a formula for the distribution of primes that would
further improve the estimate. However, this has neither been proved nor
contradicted so far.
\end{itemize}

@@

% --------------------------------------------------------------------------
\newpage
\subsubsection{Number of prime numbers in various intervals}\label{s:primhfk}
\hypertarget{primhfk}{}

@@

% --------------------------------------------------------------------------
\newpage
\subsubsection{Indexing prime numbers ($n$-th prime number)} \hypertarget{ntePrimzahl}{}\label{s:ntePrimzahl}

@@

% --------------------------------------------------------------------------
\newpage
\begin{thebibliography}{99999}
\addcontentsline{toc}{subsection}{Bibliography}

%already defined in elementaryNumberTheory.inc -> 2 davor
\bibitem[Bartholome1996]{2Bartholome1996}  \index{Bartholome 1996}
    A. Bartholomé, J. Rung, H. Kern, \\     
    {\em Zahlentheorie f\"ur Einsteiger}, Vieweg 1995, 2nd edition 1996.
